---
title: "Census analysis"
author: "Gray Collins"
date: "2025-10-17"
output: 
  html_document: 
    self_contained: true
params: 
  git_path: "C:/Users/Gray Collins/Documents/GitHub/kampala-route-census"
  drive_path: "G:/Shared drives/ugandatransit/uganda_transit_archives"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(haven)
library(openxlsx)
library(tidyverse)
library(lubridate)
library(sf)
library(tmap)
library(rmapshaper)
# library(survival)
# library(osmdata)
# library(vroom)
# library(fixest)
# library(ggfortify)
# library(viridis)
# library(spatialreg)
# library(dynlm)
# library(forecast)
library(stargazer) 
```

```{r read json data}
# Gridcell-level routes and population: 
grid_routes <- file.path(params$git_path, "kampala-geography", "grid_routes.geojson") %>% st_read()
grid_pop <- file.path(params$git_path, "population-data", "grid-pop.csv") %>% read_csv()
# grid_pop_2 <- file.path(params$git_path, "population-data", "grid-pop-2015_2024.csv") %>% read_csv()

grid <- left_join(grid_routes, grid_pop, by='id') %>% 
  rename_with(~ gsub("F", "nroutes_", .x, fixed=T), 
              starts_with("F"))
 



# Neighborhood-level routes and population: 
nhood_routes <- file.path(params$git_path, "kampala-geography", "neighborhood-routes.geojson") %>% st_read() %>% 
  # NOTE: there are multiple rows in nhood_routes for Central Division - correct: 
  mutate(name = case_when(
    name=="Central Division" & F2024==502 ~ "Central Division", 
    name=="Central Division" & F2024==25 ~ "Central Division (Mukono)", 
    name=="Central Division" & F2024==2 ~ "Central Division (Buikwe)", 
    name=="Central Division" & F2024==0 ~ "Central Division (Jinja)",
    .default = name
    )) 

nhood_pop <- file.path(params$git_path, "population-data", "nhood-pop.csv") %>% read_csv()
# nhood_pop_2 <- file.path(params$git_path, "population-data", "nhood-pop-2015_2024.csv") %>% read_csv() %>% 
#   select(neighborhood, pop_2021, pop_2022, pop_2023, pop_2024)
# 
# # Join population data 
# nhood_pop <- tidylog::full_join(nhood_pop_1, nhood_pop_2, by='neighborhood')

nhood <- tidylog::left_join(nhood_routes, nhood_pop, by=c("name"="neighborhood")) %>%
  # Create neighborhood identifier
  mutate(id = name) %>% 
  rename_with(~ gsub("F", "nroutes_", .x, fixed=T), 
              starts_with("F"))

# Routes shapefile: 
routes <- file.path(params$git_path, "route-data", "routes_final", "routes.shp") %>% st_read()

```

```{r reshape grid and neighborhood level combined data}
# Grid-level
grid_long <- grid %>%
  ungroup() %>% 
  pivot_longer(
    cols = matches("(.+)20([0-9])([0-9])"),
    names_to = c(".value", "year"),
    names_sep = "_"
  ) %>% 
  arrange(id, year) %>% 
  group_by(id) %>% 
  mutate(
    popchange_yoy = pop - lag(pop),
    popchange_pct_yoy = 100*popchange_yoy/lag(pop),   # % change in population
    routechange_yoy = nroutes - lag(nroutes),
    routes_any = ifelse(nroutes>0, 1,0)
  )

# Nhood-level
nhood_long <- nhood %>% 
  ungroup() %>% 
  pivot_longer(
    cols = matches("(.+)20([0-9])([0-9])"),
    names_to = c(".value", "year"),
    names_sep = "_"
  ) %>% 
  distinct(id, year, .keep_all=TRUE) %>% 
  arrange(id, year) %>% 
  group_by(id) %>% 
  mutate(
    popchange_yoy = pop - lag(pop),
    popchange_pct_yoy = 100*popchange_yoy/lag(pop),   # % change in population 
    routechange_yoy = nroutes - lag(nroutes),
    routes_any = ifelse(nroutes>0, 1,0)
  )

```

# Maps

## Taxi routes & population, by neighborhood 
```{r route map}
tmap_mode("plot")

tm_shape(routes %>% filter(F2019==1), is.main=FALSE) + 
    tm_lines(col='red') + 
  tm_shape(nhood_long %>% filter(year==2019) %>% select(id, pop) %>%
             mutate(id = id %>% str_replace_all("Town Council", "T.C.") %>% str_replace_all("Division", "Div."))) +
  tm_text(text = 'id', size=1) +
  tm_polygons(fill='pop', fill_alpha=0.5) + 
  # tm_shape(grid_long %>% filter(year==2019)) + 
  #   tm_polygons(hover='id', fill='pop', fill_alpha=0.5) + 
  tm_title("Taxi routes (2019) and neighborhoods") 
```


## Population

```{r population maps, resuts='asis'}
map_pop <- function(data, border_alpha, yr){
  data <- data %>% 
    select(id, starts_with("pop"))
  
  popmap <- tm_shape(data) +
   tm_polygons(fill = paste0("pop_", yr),
               col_alpha = border_alpha, 
               hover = 'id',
               fill.scale = tm_scale_continuous(
                 values = viridis::viridis(7),
                 ticks = c(100, 1000, 2000, 5000, 10000, 15000)),
               fill_alpha = .7,
               fill.legend = tm_legend(
                  title = paste0("Population - ", yr),
                  component.autoscale = FALSE
               )) + 
   tm_title(paste0("Population of Kampala - ", yr))
  
  return(popmap)
}

# map_pop(grid, border_alpha=0, yr=2000)
# map_pop(grid, border_alpha=0, yr=2005)
# map_pop(grid, border_alpha=0, yr=2010)
# map_pop(grid, border_alpha=0, yr=2015)
map_pop(grid, border_alpha=0, yr=2020)

```

## Percent change in population 

```{r pct change population maps}
map_pop_pctchange <- function(data, border_alpha, yr){
  data <- data %>% 
    filter(year==yr)
  
  scalemin <- min(data$popchange_pct_yoy, na.rm=T)
  scalemax <- max(data$popchange_pct_yoy, na.rm=T)
  
  popmap_pctchange <-  tm_shape(data) +
   tm_polygons(fill = "popchange_pct_yoy",
               col_alpha = border_alpha, 
               hover = 'id',
               fill.scale = tm_scale_continuous(
                 values = RColorBrewer::brewer.pal(9, 'RdBu'),
                 outliers.trunc = TRUE,
                 ticks = c(-25, -10, -5, -1, 0, 1, 5, 10, 25)),
               fill_alpha = .7,
               component.autoscale = FALSE,
               fill.legend = tm_legend(
                  title = paste0("Pct change in population - ", yr)
               )) + 
   tm_title(paste0("% change in population - ", yr)) + 
    tmap_options(component.autoscale = FALSE)
  
  return(popmap_pctchange)
}

# map_pop_pctchange(grid_long, border_alpha=0, yr=2005)
# map_pop_pctchange(grid_long, border_alpha=0, yr=2010)
# map_pop_pctchange(grid_long, border_alpha=0, yr=2015)
map_pop_pctchange(grid_long, border_alpha=0, yr=2020)  

```

# Plots

## Route growth versus population growth 

It seems like there is a logarithmically shaped relationship between population and route growth in neighborhoods with smaller populations. But for neighborhoods with larger populations, this relationship seems to take a different shape. 

```{r plots route x population}
# N routes X Log population 
ggplot(nhood_long, 
       aes(x=pop, y=nroutes, group=id)) + 
  # geom_point() + 
  geom_line() +
  labs(title = "Number of routes & population - all neighborhoods")


# N routes & population 
ggplot(nhood_long, 
       aes(x=popchange_yoy, y=routechange_yoy, group=id,
           size=pop)) + 
  geom_point(color='blue', alpha=0.3) + 
  labs(title = "Year-over-year route growth & absolute population growth - all neighborhoods")
```

## Population change (neighborhood-level, year-over-year)

Note: Population growth rates by neighborhood are mostly constant over the years, but oddly bunched at the 2.5% and 6% thresholds. The line graph is messy but demonstrates that neighborhoods typically experience a constant growth rate over the years. 

What might explain the bunching at 2.5% and 6%?  

```{r plots route x percent change in population}
ggplot(nhood_long, 
       aes(x=popchange_pct_yoy, y=routechange_yoy, group=id)) + 
  geom_point() + 
  labs(title = "Year-over-year route growth & percent change in population - all neighborhoods")


ggplot(nhood_long, aes(x=popchange_pct_yoy)) + 
  geom_histogram() +
  scale_x_continuous(breaks = seq(-2,15,by=1)) +
  labs(title = "% change in population - histogram")

ggplot(nhood_long, aes(x=year, y=popchange_pct_yoy, group=id, color=id)) + 
  geom_line(show.legend=F) + 
  labs(title = "% change in population by neighborhood")
 
```

Below compares neighborhood X year combinations with 0 taxi routes to neighborhood X year combinations with any routes. 

```{r}

ggplot(nhood_long, 
       aes(x=popchange_pct_yoy)) + geom_density() + 
  labs(title = "% change in population by neighborhood") +
  facet_wrap(~factor(routes_any, levels=c(0,1), labels=c("No taxi routes", "1+ taxi routes")))

ggplot(nhood_long, 
       aes(x=pop)) + geom_histogram() + 
  labs(title = "Total population by neighborhood") +
  facet_wrap(~factor(routes_any, levels=c(0,1), labels=c("No taxi routes", "1+ taxi routes")))

```

Log-population by neighborhood:

```{r scatter plots route and population by neighborhood, fig.width=8, fig.height=15}
# ggplot(data = nhood_long,
#        aes(x=year, y=nroutes, group=id)) + 
#   geom_line() + 
#   facet_wrap(~str_wrap(id, 30), nrow=12) + 
#   scale_x_discrete(breaks = seq(2000,2025, by=5)) +
#   labs(title = "Number of routes, by neighborhood")

ggplot(data = nhood_long,
       aes(x=year, y=log(pop), group=id)) + 
  geom_line() + 
  facet_wrap(~str_wrap(id, 30), nrow=12) + 
  scale_x_discrete(breaks = seq(2000,2025, by=5)) +
  labs(title = "Log population, by neighborhood")

# ggplot(data = nhood_long,
#        aes(x=year, y=pop, group=id)) + 
#   geom_line() + 
#   facet_wrap(~str_wrap(id, 30), nrow=12) + 
#   scale_x_discrete(breaks = seq(2000,2025, by=5)) +
#   labs(title = "Total population, by neighborhood")
```



# Analysis 

- Look among gridcells which are similar in population, in the ?th percentile of gridcells (find a reasonable cut point judging by population data). 
- Identify which gridcells “earned” service (based on population) and those which didn’t (those with low population, but got a route anyway).

- Service indicator: Binary, whether a gridcell has routes going through it or not
- Diff in means (population) between these at different time intervals in future
    - Time horizon: test several
- Diff: between corridors which got a route and those which didnt


## Define reference year
IMPORTANT: The code below calculates differences in population and route coverage relative to this reference year; change it before rerunning the below analysis  
```{r define base year for analysis}
yr <- 2019
```


## Find population cutpoint 

First create an indicator for whether a route was added 

```{r}

# Year in which service began 
start_yr <- grid_long %>% 
  st_drop_geometry() %>% 
  as.data.frame() %>% 
  group_by(id) %>% 
  arrange(year) %>% 
  filter(nroutes>0) %>%
  summarise(start_year = min(year))

# Value labels for the service_status indicator 
servicelabs <- c(paste0("[0] No routes in ", yr),
                 paste0("[1] Newly served in ", yr),
                 paste0("[2] Served ", yr, ", unchanged from ", yr-1),
                 paste0("[3] Served ", yr, ", increase from ", yr-1),
                 paste0("[4] Served ", yr, ", decrease from ", yr-1))

grid_long_changes <- grid_long %>%
  left_join(start_yr, by='id') %>% 
  group_by(id) %>% 
  # Indicate unserved and newly served areas  
  mutate(
    unserved = ifelse(nroutes==0, 1, 0), 
    newly_served = ifelse(year==start_year, 1, 0),
    
    service_status = case_when(
      year==yr & nroutes==0 ~ 0,
      year==yr & nroutes > 0 & lag(nroutes)==0 ~ 1,
      year==yr & nroutes==lag(nroutes) & nroutes>0 ~ 2,
      year==yr & nroutes > 0 & lag(nroutes) > 0 & nroutes>lag(nroutes) ~ 3,
      year==yr & nroutes < lag(nroutes) ~ 4) %>% as.factor(),
    service_status_l = case_when(
      year==yr & nroutes==0 ~ paste0("[0] No routes in ", yr),
      year==yr & nroutes > 0 & lag(nroutes)==0 ~ paste0("[1] Newly served in ", yr),
      year==yr & nroutes==lag(nroutes) & nroutes>0 ~ paste0("[2] Served ", yr, ", unchanged from ", yr-1),
      year==yr & nroutes > 0 & lag(nroutes) > 0 & nroutes>lag(nroutes) ~ paste0("[3] Served ", yr, ", increase from ", yr-1),
      year==yr & nroutes < lag(nroutes) ~ paste0("[4] Served ", yr, ", decrease from ", yr-1)) %>%
      as.factor()
  ) %>% 
  fill(service_status, .direction='downup') %>% 
  fill(service_status_l, .direction='downup')
 

# Summary: 
grid_long_changes %>% 
  filter(year==yr) %>% 
  group_by(service_status_l) %>% 
  summarise(n=n())
```

Density plot: Population in served vs. non-served gridcells (reference year: 2019). The second plot (showing population in only those gridcells with any changes at all) doesn't seem to suggest a clear population threshold at which a gridcell "earns" route service. Among those [1] newly served in 2019, there are too few observations to tell a cutoff; if there was any cutoff, perhaps it'd be pop=2500, but it's counter-intuitive that most gridcells are below that threshold and only n=2 are above it. Perhaps among routes with [3] increased service relative to 2018, the distribution seems almost bimodal and pop=7500 could be a cutoff for increasing service. But this doesn't explain the emergence of routes in previously unserved gridcells. 

These plots + the table above show there are very few gridcells with any service changes whatsoever.

```{r}

# Population: By service status
ggplot(grid_long_changes %>% filter(year==yr)) + 
  geom_density(aes(x=pop)) + 
  labs(title = paste0("Population (", yr, "): By route service status"),
       caption = "note - flexible axes") + 
  facet_wrap(~service_status_l)


# Population: Closer look at routes with any changes
ggplot(grid_long_changes %>% filter(year==yr & service_status!=0)) + 
  geom_histogram(aes(x=pop), bins=50) + 
  labs(title = paste0("Population (", yr, "): By route service status"),
       caption = "note - flexible axes") + 
  facet_wrap(~service_status_l)

```

## Visualize new service corridors 

Since there are so few gridcells where new service was added, can we try to classify them in another way--namely, is it visually apparent which transit corridor they are part of? 

The map below shows the gridcells which did not have any route coverage in the previous year, but do have coverage during the test year. The routes which were inactive last year but active this year are also shown in red. 

For the year 2019, there are a few clearly visible corridor additions: 
- In the West, Wakiso neighborhood, two routes divert to Kigoma, and another continues past Bukasa along Sentema Road. 
- To the Southwest, in Kyengera Town council, a new route from Kisenyi-Gwanda branches south from Kampala-Masaka Road and continues just across the Entebbe border via Busembe and Nanziga
- In the South, a new route to Entebbe operates along the Kampala-Entebbe Expressway (which was opened in 2018), adding new service to several previously unserved gridcells in Kyengera and Kajjansi. 
- There are other, minor additions to routes that typically don't result in a previously unserved gridcell getting service. 

```{r}
newservicemapdata <- grid_long_changes %>% 
  filter(year==yr & service_status==1) %>%
  select(id, neighborhood, nroutes, pop, popchange_yoy, popchange_pct_yoy, service_status_l)

# Filter routes to only those which were active this year  
routes_yr <- routes %>% 
  rename(active_now = paste0("F", paste(yr)),
         active_last_year = paste0("F", paste(yr-1))) %>%
  filter(active_now==1) %>% 
  select(route_id, stage_code, Origin, Stage, ObjectID_1, layer, active_now, active_last_year) %>% 
  mutate(hover = paste0("Route ", route_id, ", Stage ", stage_code, ", Origin ", Origin),
         color = case_when(   # note - treat values of NA in the routes dataset like 0
           active_now==1 & (active_last_year==0 | is.na(active_last_year)) ~ "blue", 
           .default = "black"))
 

# Map: 
tm_basemap("OpenStreetMap") +
tm_title(paste0("New routes and newly served gridcells: ", yr)) +
  # tm_shape(nhood) + tm_polygons(fill_alpha=0) + 
  tm_shape(newservicemapdata) + tm_polygons(fill = 'blue', fill_alpha=0.5) + 
  tm_shape(routes_yr) + tm_lines(hover = 'hover', col = 'color', col_alpha=0.5, lwd=2)
```


The below map shows gridcells where the number of routes increased (green), decreased (red), or did not change (grey) based on the prior year. Previously unserved gridcells with route service this year are still shown in blue. 

```{r, fig.width=8, fig.height=8}
# Map ALL gridcells by service status: 
allservicemapdata <- grid_long_changes %>% 
  filter(year==yr) %>% 
  mutate(fill = case_match(service_status,
                           "0" ~ NA,
                           "1" ~ "blue",
                           "2" ~ "white", 
                           "3" ~ "green",
                           "4" ~ "red"))

tm_basemap("OpenStreetMap") +
tm_title(paste0("New routes and all service changes in gridcells: ", yr)) + 
  tm_shape(allservicemapdata) + tm_polygons(fill = 'fill', fill_alpha=0.5, col=NA, col_alpha=0, lwd=0) + 
  tm_shape(routes_yr) + tm_lines(hover = 'hover', col = 'color', col_alpha=0.5, lwd=1)

```


## Compare for other years 

```{r function to get grid_long_changes for other years, echo=F}

get_long_changes <- function(refyear) {
  # Value labels for the service_status indicator 
  servicelabs <- c(paste0("[0] No routes in ", refyear),
                 paste0("[1] Newly served in ", refyear),
                 paste0("[2] Served ", refyear, ", unchanged from ", refyear-1),
                 paste0("[3] Served ", refyear, ", increase from ", refyear-1),
                 paste0("[4] Served ", refyear, ", decrease from ", refyear-1))
  
  df <- grid_long %>%
    left_join(start_yr, by='id') %>% 
    group_by(id) %>% 
    mutate(
      unserved = ifelse(nroutes==0, 1, 0), 
      newly_served = ifelse(year==refyear, 1, 0),
      service_status = case_when(
        year==refyear & nroutes==0 ~ 0,
        year==refyear & nroutes > 0 & lag(nroutes)==0 ~ 1,
        year==refyear & nroutes==lag(nroutes) & nroutes>0 ~ 2,
        year==refyear & nroutes > 0 & lag(nroutes) > 0 & nroutes>lag(nroutes) ~ 3,
        year==refyear & nroutes < lag(nroutes) ~ 4) %>% as.factor(),
      service_status_l = factor(service_status,
                                levels = c(0,1,2,3,4), 
                                labels = servicelabs)) %>% 
    fill(service_status, .direction='downup') %>% 
    fill(service_status_l, .direction='downup') 

    return(df)
}

map_long_changes <- function(refyear) {
  # Get data and add fill aesthetics: 
  df <- get_long_changes(refyear)
  
  servicemapdata <- df %>% 
    filter(year==refyear) %>% 
    mutate(fill = case_match(
      service_status,
      "0" ~ NA, "1" ~ "blue", "2" ~ "white", "3" ~ "green", "4" ~ "red")) 
  
  # Routes active this year: 
  routesyr <- routes %>% 
  rename(active_now = paste0("F", paste(refyear)),
         active_last_year = paste0("F", paste(refyear-1))) %>%
  filter(active_now==1) %>% 
  select(route_id, stage_code, Origin, Stage, ObjectID_1, layer, active_now, active_last_year) %>% 
  mutate(hover = paste0("Route ", route_id, ", Stage ", stage_code, ", Origin ", Origin),
         color = case_when(   # note - treat values of NA in the routes dataset like 0
           active_now==1 & (active_last_year==0 | is.na(active_last_year)) ~ "blue", 
           .default = "black"))
 
  # Map: 
  tm_basemap("OpenStreetMap") +
  tm_title(paste0("New routes and all service changes in gridcells: ", refyear)) +
    tm_shape(servicemapdata) + tm_polygons(fill = 'fill', fill_alpha=0.5, col=NA, col_alpha=0, lwd=0) + 
    tm_shape(routesyr) + tm_lines(hover = 'hover', col = 'color', col_alpha=0.5, lwd=2)
}

```

Maps for other years: 
Blue = New service, Green = Increased service, Red = Decreased service, White = Unchanged service, (relative to previous year) 
- 2018: No newly served gridcells 
- 2017: Several new corridors expand service into the N and NW areas of the city, with gridcells around Nansana and north of Tula receiving new routes 
- 2016: Some changes in connections but only one newly served gridcell near Naaya Estate in the northeast; most new routes are small branches or redirections
- 2015: Several extensions provide routes in newly served parts of Mukono, NE of Gayaza, Lugo, and north of Bukasa

```{r, fig.width=8, fig.height=8}
map_long_changes(2014)
map_long_changes(2015)
map_long_changes(2016)
map_long_changes(2017)
map_long_changes(2019)
```

 





