---
title: "Untitled"
author: "Gray Collins"
date: "2025-10-17"
output: html_document
params: 
  git_path: "C:/Users/Gray Collins/Documents/GitHub/kampala-route-census"
  drive_path: "G:/Shared drives/ugandatransit/uganda_transit_archives"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(openxlsx)
library(tidyverse)
library(lubridate)
library(survival)
library(osmdata)
library(vroom)
library(sf)
library(tmap)
library(fixest)
library(ggfortify)
library(viridis)
library(spatialreg)
library(dynlm)
library(forecast)
library(stargazer)
```

```{r read json data}
# Gridcell-level routes and population: 
grid_routes <- file.path(params$git_path, "kampala-geography", "grid_routes.geojson") %>% st_read()
grid_pop <- file.path(params$git_path, "population-data", "grid-pop.csv") %>% read_csv()

grid <- left_join(grid_routes, grid_pop, by='id') %>% 
  rename_with(~ gsub("F", "nroutes_", .x, fixed=T), 
              starts_with("F"))


# Neighborhood-level routes and population: 
nhood_routes <- file.path(params$git_path, "kampala-geography", "neighborhood-routes.geojson") %>% st_read() %>% 
  # NOTE: there are multiple rows in nhood_routes for Central Division - correct: 
  mutate(name = case_when(
    name=="Central Division" & F2024==502 ~ "Central Division", 
    name=="Central Division" & F2024==25 ~ "Central Division (Mukono)", 
    name=="Central Division" & F2024==2 ~ "Central Division (Buikwe)", 
    name=="Central Division" & F2024==0 ~ "Central Division (Jinja)",
    .default = name
    )) 

nhood_pop <- file.path(params$git_path, "population-data", "nhood-pop.csv") %>% read_csv()

nhood <- tidylog::left_join(nhood_routes, nhood_pop, by=c("name"="neighborhood")) %>%
  # Create neighborhood identifier
  mutate(id = name) %>% 
  rename_with(~ gsub("F", "nroutes_", .x, fixed=T), 
              starts_with("F"))

# Routes shapefile: 
routes <- file.path(params$git_path, "route-data", "routes_final", "routes.shp") %>% st_read()

```

```{r reshape grid and neighborhood level combined data}
# Grid-level
grid_long <- grid %>%
  ungroup() %>% 
  pivot_longer(
    cols = matches("(.+)20([0-9])([0-9])"),
    names_to = c(".value", "year"),
    names_sep = "_"
  ) %>% 
  arrange(id, year) %>% 
  group_by(id) %>% 
  mutate(
    popchange_yoy = pop - lag(pop),
    popchange_pct_yoy = 100*popchange_yoy/lag(pop),   # % change in population
    routechange_yoy = nroutes - lag(nroutes),
    routes_any = ifelse(nroutes>0, 1,0)
  )

# Nhood-level
nhood_long <- nhood %>% 
  ungroup() %>% 
  pivot_longer(
    cols = matches("(.+)20([0-9])([0-9])"),
    names_to = c(".value", "year"),
    names_sep = "_"
  ) %>% 
  distinct(id, year, .keep_all=TRUE) %>% 
  arrange(id, year) %>% 
  group_by(id) %>% 
  mutate(
    popchange_yoy = pop - lag(pop),
    popchange_pct_yoy = 100*popchange_yoy/lag(pop),   # % change in population 
    routechange_yoy = nroutes - lag(nroutes),
    routes_any = ifelse(nroutes>0, 1,0)
  )

```

# Maps

## Taxi routes & population, by neighborhood 
```{r route map}
tmap_mode("view")

tm_shape(routes %>% filter(F2019==1), is.main=FALSE) + 
    tm_lines(col='red') + 
  tm_shape(nhood_long %>% filter(year==2019) %>% select(id, pop) %>%
             mutate(id = id %>% str_replace_all("Town Council", "T.C.") %>% str_replace_all("Division", "Div."))) +
  tm_shape(grid_long %>% filter(year==2019)) + 
    tm_polygons(hover='id', fill='pop', fill_alpha=0.5) + 
    tm_text(text = 'id', size=1) + 
  tm_title("Taxi routes (2019) and neighborhoods") 
```


## Population

```{r population maps}
map_pop <- function(data, border_alpha, yr){
  data <- data %>% 
    select(id, starts_with("pop"))
  
  # tm_basemap("OpenStreetMaps") + 
  popmap <- tm_shape(data) +
   tm_polygons(fill = paste0("pop_", yr),
               col_alpha = border_alpha, 
               hover = 'id',
               fill.scale = tm_scale_continuous(
                 values = viridis::viridis(7),
                 ticks = c(100, 1000, 2000, 5000, 10000, 15000)),
               fill_alpha = .7,
               fill.legend = tm_legend(
                  title = paste0("Population - ", yr),
                  component.autoscale = FALSE
               )) + 
   tm_title(paste0("Population of Kampala - ", yr))
}

map_pop(grid, border_alpha=0, yr=2000)
map_pop(grid, border_alpha=0, yr=2005)
map_pop(grid, border_alpha=0, yr=2010)
map_pop(grid, border_alpha=0, yr=2015)
map_pop(grid, border_alpha=0, yr=2019)

```

## Percent change in population 

```{r pct change population maps}
map_pop_pctchange <- function(data, border_alpha, yr){
  data <- data %>% 
    filter(year==yr)
  
  scalemin <- min(data$popchange_pct_yoy, na.rm=T)
  scalemax <- max(data$popchange_pct_yoy, na.rm=T)
  
  tm_basemap("OpenStreetMaps") + 
    tm_shape(data) +
   tm_polygons(fill = "popchange_pct_yoy",
               col_alpha = border_alpha, 
               hover = 'id',
               fill.scale = tm_scale_continuous(
                 values = RColorBrewer::brewer.pal(9, 'RdBu'),
                 outliers.trunc = TRUE,
                 ticks = c(-25, -10, -5, -1, 0, 1, 5, 10, 25)),
               fill_alpha = .7,
               component.autoscale = FALSE,
               fill.legend = tm_legend(
                  title = paste0("Pct change in population - ", yr)
               )) + 
   tm_title(paste0("Population of Kampala - ", yr)) + 
    tmap_options(component.autoscale = FALSE)
}

map_pop_pctchange(grid_long, border_alpha=0, yr=2005)
map_pop_pctchange(grid_long, border_alpha=0, yr=2010)
map_pop_pctchange(grid_long, border_alpha=0, yr=2015)
map_pop_pctchange(grid_long, border_alpha=0, yr=2020) 

```

# Plots

## Route growth versus population growth 

It seems like there is a logarithmically shaped relationship between population and route growth in neighborhoods with smaller populations. But for neighborhoods with larger populations, this relationship seems to take a different shape. 

```{r plots route x population}
# N routes X Log population 
ggplot(nhood_long, 
       aes(x=pop, y=nroutes, group=id)) + 
  # geom_point() + 
  geom_line() +
  labs(title = "Number of routes & population - all neighborhoods")


# N routes & population 
ggplot(nhood_long, 
       aes(x=popchange_yoy, y=routechange_yoy, group=id,
           size=pop)) + 
  geom_point(color='blue', alpha=0.3) + 
  labs(title = "Year-over-year route growth & absolute population growth - all neighborhoods")
```


Note: Population growth rates by neighborhood are mostly constant over time, but oddly bunched at the 2.5% and 6% thresholds. Is this an artifact of the data somehow? 

```{r plots route x percent change in population}
ggplot(nhood_long, 
       aes(x=popchange_pct_yoy, y=routechange_yoy, group=id)) + 
  geom_point() + 
  labs(title = "Year-over-year route growth & percent change in population - all neighborhoods")


ggplot(nhood_long, aes(x=popchange_pct_yoy)) + 
  geom_histogram() +
  scale_x_continuous(breaks = seq(-2,15,by=1)) +
  labs(title = "% change in population - histogram")
```

Below compares neighborhood X year combinations with 0 taxi routes to neighborhood X year combinations with any routes. Interestingly 

```{r}
ggplot(nhood_long, aes(x=year, y=popchange_pct_yoy, group=id, color=id)) + 
  geom_line(show.legend=F) + 
  labs(title = "% change in population by neighborhood")
 
ggplot(nhood_long, 
       aes(x=popchange_pct_yoy)) + geom_density() + 
  labs(title = "% change in population by neighborhood") +
  facet_wrap(~factor(routes_any, levels=c(0,1), labels=c("No taxi routes", "1+ taxi routes")))

ggplot(nhood_long, 
       aes(x=pop)) + geom_histogram() + 
  labs(title = "Total population by neighborhood") +
  facet_wrap(~factor(routes_any, levels=c(0,1), labels=c("No taxi routes", "1+ taxi routes")))

```


```{r scatter plots route and population by neighborhood, fig.width=8, fig.height=15}
ggplot(data = nhood_long,
       aes(x=year, y=nroutes, group=id)) + 
  geom_line() + 
  facet_wrap(~str_wrap(id, 30), nrow=12) + 
  scale_x_discrete(breaks = seq(2000,2025, by=5)) +
  labs(title = "Number of routes, by neighborhood")

ggplot(data = nhood_long,
       aes(x=year, y=log(pop), group=id)) + 
  geom_line() + 
  facet_wrap(~str_wrap(id, 30), nrow=12) + 
  scale_x_discrete(breaks = seq(2000,2025, by=5)) +
  labs(title = "Log population, by neighborhood")
```



# Analysis 

Begin with the year 2024. 
- Look among gridcells which are similar in population, in the ?th percentile of gridcells (find a reasonable cut point judging by population data). 
- Identify which gridcells “earned” service (based on population) and those which didn’t (those with low population, but got a route anyway).

- Service indicator: Binary, whether a gridcell has routes going through it or not
- Diff in means (population) between these at different time intervals in future
    - Time horizon: test several
- Diff: between corridors which got a route and those which didnt


## Find population cutpoint 

First create an indicator for whether a route was added 

```{r}

# Base year for analysis 
yr <- 2019


# Year in which service began 
start_yr <- grid_long %>% 
  st_drop_geometry() %>% 
  as.data.frame() %>% 
  group_by(id) %>% 
  arrange(year) %>% 
  filter(nroutes>0) %>%
  summarise(start_year = min(year))

# Value labels for the service_status indicator 
servicelabs <- c(paste0("[0] No routes in ", yr),
                 paste0("[1] Newly served in ", yr),
                 paste0("[2] Served ", yr, ", unchanged from ", yr-1),
                 paste0("[3] Served ", yr, ", increase from ", yr-1),
                 paste0("[4] Served ", yr, ", decrease from ", yr-1))

# Service indicator 
grid_long_changes <- grid_long %>%
  left_join(start_yr, by='id') %>% 
  group_by(id) %>% 
  # Indicate unserved and newly served areas  
  mutate(
    unserved = ifelse(nroutes==0, 1, 0), 
    newly_served = ifelse(year==start_year, 1, 0),
    
    service_status = case_when(
      year==yr & nroutes==0 ~ 0,
      year==yr & nroutes > 0 & lag(nroutes)==0 ~ 1,
      year==yr & nroutes==lag(nroutes) & nroutes>0 ~ 2,
      year==yr & nroutes > 0 & lag(nroutes) > 0 & nroutes>lag(nroutes) ~ 3,
      year==yr & nroutes < lag(nroutes) ~ 4) %>% as.factor(),
    service_status_l = case_when(
      year==yr & nroutes==0 ~ paste0("[0] No routes in ", yr),
      year==yr & nroutes > 0 & lag(nroutes)==0 ~ paste0("[1] Newly served in ", yr),
      year==yr & nroutes==lag(nroutes) & nroutes>0 ~ paste0("[2] Served ", yr, ", unchanged from ", yr-1),
      year==yr & nroutes > 0 & lag(nroutes) > 0 & nroutes>lag(nroutes) ~ paste0("[3] Served ", yr, ", increase from ", yr-1),
      year==yr & nroutes < lag(nroutes) ~ paste0("[4] Served ", yr, ", decrease from ", yr-1)) %>%
      as.factor()
  ) %>% 
  fill(service_status, .direction='downup') %>% 
  fill(service_status_l, .direction='downup')
 

# Summary: 
grid_long_changes %>% 
  filter(year==yr) %>% 
  group_by(service_status) %>% 
  summarise(n=n())
```

Density plot: Population in served vs. non-served gridcells. The 2019 data suggests there is a cutoff around 2250 people. 

```{r}

# Population: All gridcells 
ggplot(grid_long_changes %>% filter(year==yr)) + 
  geom_density(aes(x=pop)) + 
  labs(title = paste0("Population (", yr, "): All gridcells"))


# Population: By service status 
ggplot(grid_long_changes %>% filter(year==yr)) + 
  geom_density(aes(x=pop)) + 
  labs(title = paste0("Population (", yr, "): By route service status"),
       caption = "note - flexible axes") + 
  facet_wrap(~service_status, scales='free')

```

## Identify 'treatment' areas

"Treatment" gridcells = Cells which are below the population cutoff but received route service 
"Control" gridcells = Cells which are below the population cutoff and did not receive route service 

For 2019, the population cutoff is ~ 2250 

```{r test for 2019}
yr.test <- 2019 
pop.cut <- 2250 


grid_long_changes %>% 
  filter(service_status==paste0("[0] No routes in ", yr) | 
           service_status == paste0("[1] Newly served in ", yr)) %>% 
  mutate(
    treat = ifelse(pop<pop.cut & service_status = )
  )

```




## Test differences in population growth 

Test differences in population growth for gridcells where service was added vs gridcells with constant service 

For year = 2019: Test diff in 2020 population 

```{r, fig.height=10, fig.widgth=6}
# Filter to only areas where routes were added in focus yr
grid_long_test <- grid_long_changes %>% 
  filter(service_status==paste0("[0] No routes in ", yr) | 
           service_status == paste0("[1] Newly served in ", yr))


# Test



# Plot 
ggplot(grid_long_test %>% filter(!is.na(pop))) + 
  geom_density(aes(x=pop)) + 
  facet_grid(cols=vars(service_status), rows=vars(year)) + 
  labs(title = paste0("Population by year: by service status in ", yr))

# Test association 
ols_res <- lm(popchange_pct_yoy ~ service_status + popchange_pct_yoy + lag(popchange_pct_yoy),
   data = grid_long_test) 

stargazer(ols_res, type='text')

```

