---
title: "Example Analysis"
format: html
editor: visual
---

This is an example of some initial analysis comparing the route census to the population distribution in Kampala.

## Required libraries

This example uses the following libraries

```{r, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(sf)
```

## Load data

We have four data files.

-   **grid_routes.geojson:** A spatial data file with the number of routes crossing each grid cell for each year between 2000 and 2024 (variables labeled as F{year}). There is also a unique ID number for each grid cell and a variable indicating which neighborhood it's in. This file includes the geometry of the grid cell boundaries

-   **neighborhood-routes.geojson:** A spatial data file with the number of routes crossing each neighborhood for each year between 2000 and 2024 (variables labeled as F{year}). This file includes the geometry of the neighborhood boundaries

-   **grid-pop.csv:** The estimated population of each grid cell for each year from 2000 to 2020. This data can be joined to the grid_routes layer by the grid cell ID numbers.

-   **nhood-pop.csv:** The estimated population of each neighborhood for each year from 2000 to 2020. This data can be joined to the neighborhood routes layer by the neighborhood name.

I am not sure why we're missing population data for 863 grid cells or why three neighborhood names are duplicated.

```{r, message=FALSE, results='hide'}

grid_routes <- here("kampala-geography",
                    "grid_routes.geojson") |>
  st_read()

nhood_routes <- here("kampala-geography",
                     "neighborhood-routes.geojson") |>
  st_read()

grid_pop <- here("population-data",
                 "grid-pop.csv") |>
  read_csv()

nhood_pop <- here("population-data",
                  "nhood-pop.csv") |>
  read_csv() |>
  rename(name = neighborhood)

grid <- inner_join(grid_routes, grid_pop)

nhood <- inner_join(nhood_routes, nhood_pop)
```

## Map population and route distribution

Here is a map showing the distribution of taxi routes across grid cells in 2010.

```{r, warning=FALSE}

ggplot(grid) +
  geom_sf(aes(fill = F2010),
          color = NA) + 
  scale_fill_viridis_c(name = "Number of taxi routes",
                       transform = "log",
                       breaks = 2^seq(0, 8, by=1)) +
  theme_void()
```

And here is a map showing the population variation across grid-cells in 2010:

```{r}

ggplot(grid) +
  geom_sf(aes(fill = pop_2010),
          color = NA) + 
  scale_fill_viridis_c(name = "Population",
                       transform = "log",
                       breaks = breaks <- 10^seq(0, 4, by=1),
                       labels = prettyNum(breaks, big.mark = ",")) +
  theme_void()
```

Here is a map showing the presence of routes across neighborhoods:

```{r, warning=FALSE}

ggplot(nhood) +
  geom_sf(aes(fill = F2010),
          color = NA) + 
  scale_fill_viridis_c(name = "Number of taxi routes",
                       transform = "log",
                       breaks = 2^seq(0, 8, by=1)) +
  theme_void()
```

And here is a map showing the variation on population by neighborhood in 2010.

```{r}

ggplot(nhood) +
  geom_sf(aes(fill = pop_2010),
          color = NA) + 
  scale_fill_viridis_c(name = "Population",
                       transform = "log",
                       breaks = breaks <- 10000*2^seq(0, 6, by=1),
                       labels = formatC(breaks, 
                                        format = "d",
                                        big.mark = ",")) +
  theme_void()
```

## Map changes in population and service

Here are the changes in route presence between 2009 and 2010, at the grid cell level.

```{r}

grid_change <- grid |>
  mutate(route_change = F2010 - F2009)

ggplot(grid_change) +
  geom_sf(data = nhood,
          fill = "white", color = "gray") +
  geom_sf(aes(fill = route_change),
          color = NA) +
    scale_fill_gradient2(name = "Change in\nnumber of\nroutes serving\nlocation") +
  theme_void()
```

Here are the changes in population between 2009 and 2010, at the grid cell level.

```{r}

grid_change <- grid_change |>
  mutate(pop_change = pop_2010 - pop_2009) 

ggplot(grid_change) +
  geom_sf(data = nhood,
          fill = "white", color = "gray") +
  geom_sf(aes(fill = pop_change),
          color = NA,
          alpha = 0.7) +
  scale_fill_gradient2(name = "Population\nchange") +
  theme_void()
```

Here are the changes in service at the neighborhood level between 2009 and 2010.

```{r}

nhood_change <- nhood |>
  mutate(route_change = F2010 - F2009)

ggplot(nhood_change) +
  geom_sf(aes(fill = route_change),
          color = "gray") +
  scale_fill_gradient2(name = "Change in\nnumber of\nroutes serving\narea") +
  theme_void()
```

And here are the corresponding changes in population.

```{r}

nhood_change <- nhood_change |>
  mutate(pop_change = pop_2010 - pop_2009) 

ggplot(nhood_change) +
  geom_sf(aes(fill = pop_change),
          color = "gray") +
  scale_fill_gradient2(name = "Change in\npopulation") +
  theme_void()
```

## Comparing service changes and population changes

Here is a scatter plot comparing service changes to population changes between 2009 and 2010 at the grid-cell level.

```{r, message=FALSE}

ggplot(grid_change,
       aes(x = pop_change,
                 y = route_change)) +
  geom_point(alpha = 0.2) +
  stat_smooth(method = "lm") +
  scale_x_continuous(name = "Change in population between 2009 and 2010") +
  scale_y_continuous(name = "Change in number of routes between 2009 and 2010") +
  theme_minimal()
```

Here is a similar plot at the neighborhood level.

```{r, message=FALSE}

ggplot(nhood_change,
       aes(x = pop_change,
                 y = route_change)) +
  geom_point(alpha = 0.2) +
  stat_smooth(method = "lm") +
  scale_x_continuous(name = "Change in population between 2009 and 2010") +
  scale_y_continuous(name = "Change in number of routes between 2009 and 2010") +
  theme_minimal()
```
